---
title: "MLB Player Performance and Salary"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

|     This report analyzes the relationship between Major League Baseball hitting performance 
|     and salary. The data was scrubed from various sports statistics websites, most notably 
|     sports-reference.com and ESPN. The analysis seeks to create visualizations that display 
|     clear insights within the dataset, as well as build an effective machine learning model
|     to predict a player's salary.  The data uses statistics from the 2016 to 2019 seasons.

<br>

<h2>Visualizations and Insights</h2>

***

```{r mlb1, include=FALSE}

library(rvest)
library(dplyr)
library(tidyverse)
library(stringr)
library(ggplot2)
library(plotly)
library(caret)
library(rpart)
library(rpart.plot)
library(DMwR)
library(hrbrthemes)
library(patchwork)
library(scales)
library(flextable)
library(extrafont)
library(knitr)

urlfile<-'https://raw.githubusercontent.com/smazurski/MLB/master/MLB_Batters.csv'

# Salaries of all hitters from 2016-2019 seasons with over 100 Plate Appearances (opportunities)
hitters <- read_csv(urlfile)

# Take a peek at the data
glimpse(hitters)



```


```{r mlb2, include=FALSE}

#######            #######
    #Regression Model
#######            #######

# Convert data types
hitters$Acquired <- as.factor(hitters$Acquired)
hitters$POS <- as.factor(hitters$POS)
hitters$Lg <- as.factor(hitters$Lg)

summary(hitters)
```


```{r mlb3, include=FALSE}

# Filter out Name, Team and salaries=0
hitters_model_data <- hitters %>%
  select(-Rk, -Name, -Tm) %>%
  filter(Salary > 0)

# Create Model
hitters_model <- lm(hitters_model_data$Salary~., data = hitters_model_data)
summary(hitters_model) # R2=.4997, F-stat: 45.25 

```


```{r mlb4, include=FALSE}

# Correlated Features
library(corrplot)

hitters_model_data %>%
  keep(is.numeric) %>%
  cor() %>%
  corrplot()


#<br>
#<br>
#<br>
#<h4><li>There are definitely some mutlicollinear features.</li></h4>
#<h4><li>There are many features correlated with Salary.</li></h4>
#<h4><li>Which features are the most important?</li></h4>
#<br>
#<br>


```

<h4>Importance By Feature</h4>

```{r mlb5, echo=FALSE}

# Variable Importance
imp <- varImp(hitters_model, scale=FALSE)
imp <- imp %>%
  arrange(desc(Overall))


ggplot2::ggplot(imp, aes(x=reorder(rownames(imp),Overall), y=Overall)) +
geom_point( color="blue", size=4, alpha=0.6)+
geom_segment( aes(x=rownames(imp), xend=rownames(imp), y=0, yend=Overall), 
color='skyblue') +
xlab('Variable')+
ylab('Overall Importance')+
theme_light() +
coord_flip() 

```

<h5>Insights:</h5>
<h6><li>National League, SLG, OBP and OPS are surprisingly the lowest predictors of salary.</li></h6>
<h6><li>Age, Games Played, and Intentional Walks are the highest.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

```{r mlb6, include=FALSE}

hitters_model_data <- hitters_model_data %>%
  select(-OPS, -OBP, -SLG, -R, -Yr, -RBI, -H, -BA, -`OPS+`)


hitters_model <- lm(hitters_model_data$Salary~., data = hitters_model_data)
summary(hitters_model)


#R2=.4992, F-stat: 56.19, decent improvement with the F-statistic. 


```



```{r mlb7, include=FALSE}
imp <- varImp(hitters_model, scale=FALSE)

imp <- imp %>%
  arrange(desc(Overall))

ggplot2::ggplot(imp, aes(x=reorder(rownames(imp),Overall), y=Overall)) +
geom_point( color="blue", size=4, alpha=0.6)+
geom_segment( aes(x=rownames(imp), xend=rownames(imp), y=0, yend=Overall), 
color='skyblue') +
xlab('Variable')+
ylab('Overall Importance')+
theme_light() +
coord_flip()
```


```{r mlb8, include=FALSE}

# Remove additional fields
hitters_model_data <- hitters_model_data %>%
  select(-TB, -Lg, -HR)


hitters_model <- lm(hitters_model_data$Salary~., data = hitters_model_data)
summary(hitters_model)

#<h3>Insights:</h3>
#<h4><li>R2=.4997, F-stat: 64.21</li></h4>
#<h4><li>Our model only accounts for ~50% of the variance in salary.</li></h4>
#<br>


```


<h4>Position Group Salaries and Performance</h4>

```{r mlb9, echo=FALSE}

#cor(hitters_model_data$WAR,hitters_model_data$Salary) 
# Wins Above Replacement is not very correlative with Salary

options(dplyr.summarise.inform = FALSE)

hitters_by_position <- hitters_model_data %>%
  group_by(POS) %>%
  summarise(Players=n(), 
            Average_WAR=round(mean(WAR),2),
            Average_Age=round(mean(Age),2), 
            Average_Salary=round(mean(Salary),2), 
            Median_Salary=round(median(Salary),2))

kable(hitters_by_position)

```

<h5>Insights:</h5>
<h6><li>Third Basemen have the highest Wins-Above-Replacement and Average Salaries.</li></h6>
<h6><li>Catchers are the oldest average age group.</li></h6>
<h6><li>Utility players (players who play multiple positions, IF and OF) are the largest player group.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

<h4>Median Salary and OPS by Age</h4>
```{r mlb10, echo=FALSE}

options(dplyr.summarise.inform = FALSE)

hitters_table <- hitters %>%
  group_by(Age) %>%
  summarise(players=n(), 
            AvgWAR=mean(WAR), 
            MedWar=median(WAR), 
            AvgSal=mean(Salary), 
            medSal=median(Salary),
            MedOPS=median(OPS),
            AvgOPS=mean(OPS))




p1 <- ggplot(hitters_table, aes(x=Age, y=medSal)) +
  geom_bar(color = "darkblue",  stat = "identity") +
  theme_ipsum() +
  scale_y_continuous(labels = comma, name = "Median Salary")



p2 <- ggplot(hitters_table, aes(x=Age, y=MedOPS)) +
  geom_bar(color = "darkblue",  stat = "identity") +
  theme_ipsum()


p1+p2



```


<h5>Insights:</h5>
<h6><li>If you're still playing baseball after age 27, you are very rich.</li></h6>
<h6><li>There is a clear drop-off in hitting performance as players get older.</li></h6>
<h6><li>Players aged 19-21 who make the league are highly productive hitters, but underpaid.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

<h4>Average MLB Age</h4>

```{r mlb11, echo=FALSE}

options(dplyr.summarise.inform = FALSE)

position_table <- hitters_model_data %>%
  group_by(Age, POS) %>%
  summarise(players=n(), AvgWAR=mean(WAR), MedWar=median(WAR), AvgSal=mean(Salary), medSal=median(Salary))


ggplot(hitters_table, aes(x=Age, y=players)) +
  geom_bar(stat = "identity", color = "darkblue", fill = "darkgray", size = 1.5) +
  theme_ipsum()

#median(hitters_model_data$Age) # Median age is 28
#mean(hitters_model_data$Age) # Average age is 28.57


```


<h5>Insights:</h5>
<h6><li>The average age of a player in the MLB is 28.</li></h6>
<h6><li>The distribution is skewed towards older players.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

<h4>MLB Age Distribution by Position</h4>

```{r mlb12, echo=FALSE}

hitters_age_table <- hitters_model_data %>%
  group_by(Age, POS) %>%
  summarise(players=n(), AvgWAR=mean(WAR), MedWar=median(WAR), AvgSal=mean(Salary), medSal=median(Salary))


# Age Distributions by Position
#ggplot(hitters_age_table, aes(x=Age, y=players, group = POS)) +
#  geom_bar(stat = "identity", color = "darkblue", fill = "darkgray", size = 1.5) +
#  theme_ipsum() +
#  facet_wrap(~POS)


ggplot(hitters_model_data, aes(x=Age, fill=POS)) + 
  geom_density(alpha=0.8, color = 'Black' ) +
  theme_ipsum() +
  facet_wrap(~POS)

```

<h5>Insights:</h5>
<h6><li>Middle Infielders are the youngest position group.</li></h6>
<h6><li>Catchers are the oldest position group.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

<h4>Average WAR By Position and Age</h4>

```{r mlb13, echo=FALSE}

ggplot(position_table, aes(x=Age, y=AvgWAR)) +
  geom_area(size = 1, color="black", fill="darkblue") +
  xlab("Age") +
  ylab("Average WAR") +
  theme_ipsum() +
  facet_wrap(~POS)

```


<h5>Insights:</h5>
<h6><li>First Basemen have the shortest careers.</li></h6>
<h6><li>Young Outfielders are highly valuable players.</li></h6>
<h6><li>WAR generally drops off as players get older.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

<h4>Median Salary By Acquisition Type</h4>

```{r mlb14, echo=FALSE}

options(dplyr.summarise.inform = FALSE)

# View By Acquisition Type
acquired_table <- hitters_model_data %>%
  group_by(Acquired) %>%
  summarise(players=n(), AvgWAR=mean(WAR), MedWar=median(WAR), AvgSal=mean(Salary), medSal=median(Salary)) %>%
  arrange(desc(medSal))

acquired_table %>%
  arrange(desc(medSal)) %>%
  ggplot() +
  geom_bar(aes(reorder(Acquired, medSal), medSal), stat="identity", color = "darkblue", fill = "darkgray", size=1) +
  scale_y_continuous(labels = comma, name = "Median Salary") +
  theme_ipsum() +
  coord_flip() +
  xlab("Acquisition Type")


```


<h5>Insights:</h5>
<h6><li>Players acquired through free agency are paid the highest.</li></h6>
<h6><li>Amateur players who initially make the league are paid the least.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

<h4>MLB Salary Decision Tree</h4>


```{r mlb15, echo=FALSE}

#######            #######
    # Decision Tree #
#######            #######

tree_mod <-
  rpart(
    Salary ~ .,
    method = "anova",
    data = hitters_model_data,
    control=rpart.control(maxdepth = 3)
  )


# Plot decision tree
rpart.plot(tree_mod, type = 4, extra = 101,  digits = -3)

```

<h5>Insights:</h5>
<h6><li>The node with the highest entropy is Age < 30, which tells me that players older than 30 are expensive.</li></h6>
<h6><li>The next highest nodes are AB's and Intentional Walks. This tells me that players who have more batting opportunities are paid more.</li></h6>
<h6><li>Players who are Intentionally Walked more often are usually dangerous hitters for opposing teams, which equates to a larger salary.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

<h4>Performance By Acquisition Type Table</h4>


```{r mlb16, echo=FALSE}
options(dplyr.summarise.inform = FALSE)

free_agents <- hitters_model_data %>%
  group_by(Acquired) %>%
  summarise(Players=n(),
            Average_Age=mean(Age), 
            Average_Salary=mean(Salary), 
            Median_Salary=median(Salary), 
            Average_WAR=mean(WAR), 
            Plate_Appearances=mean(PA))

kable(free_agents)

```
<h5>Insights:</h5>
<h6><li>The average free agent is 31.8 years old, has a .818 WAR, and makes $4 million.</li></h6>
<h6><li>Amateur players are the youngest, cheapest, and most valuable hitters.</li></h6>
<h6><li>Most players are either drafted, picked up as free agents, or traded.</li></h6>

<br>
<br>
<br>

\newpage

<br>
<br>

<h4>Average Salary Per Team</h4>


```{r mlb17, echo=FALSE}

options(dplyr.summarise.inform = FALSE)

## Does not include Pitchers ##
hitters_2019 <- hitters %>%
   filter(Yr=='2019')


teams_2019 <- hitters_2019 %>%
   group_by(Team=Tm) %>%
   summarise(TotalHittersSalary=sum(Salary),
             AvgSalary=mean(Salary),
             MedianSalary=median(Salary),
             AvgAge=median(Age),
             Players=n()) %>% 
  arrange(desc(AvgSalary))


team_salary <- teams_2019 %>%
   arrange(desc(AvgSalary)) %>%
   ggplot() +
   geom_bar(aes(reorder(Team, AvgSalary), AvgSalary), stat="identity", color = "darkblue", fill = "darkgray", size=1) +
   scale_y_continuous(labels = comma, name = "Average Salary") +
   theme_ipsum() +
   coord_flip() +
   xlab("Team")

team_salary


#team_age <- teams_2019 %>%
#   arrange(desc(AvgSalary)) %>%
#   ggplot() +
#   geom_bar(aes(reorder(Team, AvgSalary), AvgAge), stat="identity", color = "darkblue", fill = "darkgray", size=1) +
#   scale_y_continuous(labels = comma, name = "Average Age") +
#   theme_ipsum() +
#   xlab("Team")
#team_age

```
<h5>Insights:</h5>
<h6><li>The Los Angeles Angels have the highest average salary for position players.</li></h6>
<h6><li>The top 4 highest paying teams are west of the Mississippi River.</li></h6>

\newpage

<br>
<br>

<h4>Average Salary By Team</h4>


```{r mlb18, echo=FALSE}
options(dplyr.summarise.inform = FALSE)

ggplot(teams_2019, aes(x=AvgSalary, y=AvgAge, color=Team)) + 
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  theme_ipsum() +
  ggtitle(print(paste("Team Age vs. Salary Correlation: ", round(cor(teams_2019$AvgSalary, teams_2019$AvgAge),2))))

```

<h5>Insights:</h5>
<h6><li>The positive correlation of .56 tells me that the teams with the highest average salaries are also older.</li></h6>

<br>
<br>


<!-- ```{r mlb17} -->



<!-- plot_ly(hitters_2019, x = ~WAR, y = ~OPS, type = 'scatter', mode = 'markers', size = ~Salary, color = ~POS, colors = 'Paired', -->
<!--         sizes = c(5, 50), -->
<!--         marker = list(opacity = 0.5, sizemode = 'diameter'), -->
<!--         text = ~paste('Name:', Name, '<br>Age:', Age, '<br>Salary:', Salary)) -->


<!-- ``` -->




<!-- ```{r mlb18} -->

<!-- plot_ly(hitters_2019, x = ~WAR, y = ~OPS, type = 'scatter', mode = 'markers', size = ~Salary, color = ~Tm, colors = 'Paired', -->
<!--         sizes = c(5, 50), -->
<!--         marker = list(opacity = 0.5, sizemode = 'diameter'), -->
<!--         text = ~paste('Name:', Name, '<br>Age:', Age, '<br>Salary:', Salary)) -->



<!-- ``` -->





<!-- ```{r pressure, echo=FALSE} -->

<!-- plot_ly(hitters_2019, x = ~WAR, y = ~OPS, type = 'scatter', mode = 'markers', size = ~Salary, color = ~Acquired, colors = 'Paired', -->
<!--         sizes = c(5, 50), -->
<!--         marker = list(opacity = 0.5, sizemode = 'diameter'), -->
<!--         text = ~paste('Name:', Name, '<br>Age:', Age, '<br>Salary:', Salary)) -->




<!-- ``` -->






